---
title: "Methylation analysis"
date: "`r format(Sys.time(), '%Y %B %d')`"
output:
    html_notebook:
        self_contained: true
        df_print: paged
        fig_height: 6
        fig_width: 6
        highlight: zenburn
        theme: cosmo
        number_sections: yes
        toc: yes
        toc_depth: 6
        toc_float:
            collapsed: false
---

```{r setup}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
```

```{r}
library(ggplot2)
library(tidyverse)
library(RColorBrewer)
library(data.table)
library(ggrepel)
library(pheatmap)
library(ggpubr)
library(pracma)
library(robustHD)
library(reshape2)
library(wordspace)
library(biomaRt)
library(openxlsx)
library(readxl)
library(magrittr)
library(purrr)
```

# Read in methylation matrix and metadata

```{r}
X <- read.table('C:/Users/protti/Desktop/Methylation/1. MSEPM/methylation_matrix_178samples.txt', header = T, row.names = 1)
metadata <- read.table('C:/Users/protti/Desktop/Methylation/1. MSEPM/Tables/metadata.txt', header = T, sep='\t', row.names = 1)
```

# Calulate factor predicted values

```{r}
Y <- as.data.frame(scale(metadata, center = FALSE, scale = TRUE))
Y$added_col <- 1
```

```{r}
Ypred <- list()

for (i in 1:178){

    Xtrain = as.matrix(X[-i, ])
    Xtest = as.matrix(X[i, ])

    Ytrain = as.matrix(Y[-i, ])
    
    C = pinv(Ytrain) %*% Xtrain
    Ypred[[i]] = Xtest %*% pinv(C)
}

Ypred <- as.data.frame(do.call(rbind, Ypred))
```

```{r}
Y <- Y[,1:28]
names(Y)[c(6:11,22,28)] <- c('% Blood Neutrophils','Neutrophil count','% Blood Lymphocytes','Lymphocyte count','% Blood Monocytes','% Blood Eosinophils','Total Cholesterol','% Epithelial cells')

Ypred <- Ypred[,1:28]
colnames(Ypred) <- colnames(Y)
```

```{r}
cormat <- as.data.frame(round(cor(Ypred,Y, method='spearman'),2))
```

```{r}
ord <- hclust(dist(cormat, method = "euclidean"), method = "complete" )$order
cormat <- as.matrix(cormat[rownames(cormat)[ord],rownames(cormat)[ord] ])

melted_cormat <- melt(cormat)
melted_cormat$diag <- melted_cormat$Var1 == melted_cormat$Var2
melted_cormat$diag[!melted_cormat$diag] <- NA
```

```{r}
ggplot(data = melted_cormat, aes(x=Var2, y=Var1, fill=value)) + 
  geom_tile(color = "darkgrey")+
  geom_tile(data = melted_cormat[!is.na(melted_cormat$diag), ], aes(color = diag), size = 0.5) +
  scale_color_manual(guide = FALSE, values = c(`TRUE` = "black"))+
  scale_fill_gradient2(high = "#FDE725FF", low = "#440154FF", mid = "white", midpoint = 0, limit = c(-1,1), space = "Lab", name="Spearman\nCorrelation") +
  geom_text(aes(Var2, Var1, label = value), color = "black", size = 2.5) +
theme(axis.title = element_blank(), 
  panel.background = element_blank(), axis.text.x = element_text(angle=45, hjust=1,vjust=1), axis.text=element_text(color='black', size=8))
```


# Keep factors with correlation > 0.3

```{r}
factors <- melted_cormat[melted_cormat$diag == 'TRUE' & melted_cormat$value > 0.3, ] 
factors <- drop_na(factors)
factors$Var1
```
Remove % lymphocytes, Lymphocyte count, CRP, NLR (redundant factors).

# Re-run predictions only for the 9 factors

```{r}
metadata <- metadata[,c(1,2,6,12,13,16,21,27,28)]
Y <- as.data.frame(scale(metadata, center = FALSE, scale = TRUE))
Y$added_col <- 1
```

```{r}
Ypred <- list()

for (i in 1:178){

    Xtrain = as.matrix(X[-i, ])
    Xtest = as.matrix(X[i, ])

    Ytrain = as.matrix(Y[-i, ])
    
    C = pinv(Ytrain) %*% Xtrain
    Ypred[[i]] = Xtest %*% pinv(C)
}
Ypred <- as.data.frame(do.call(rbind, Ypred))
```

```{r}
Y <- Y[,1:9]
Ypred <- Ypred[,1:9]

cormat <- as.data.frame(round(cor(Ypred,Y, method='spearman'),2))
names(cormat)[c(3,9)] <- c('% Blood Neutrophils','% Epithelial cells')
rownames(cormat) <- colnames(cormat)

cormat <- as.matrix(cormat)
ord <- hclust(dist(cormat, method = "euclidean"), method = "complete")$order
cormat <- cormat[rownames(cormat)[ord],rownames(cormat)[ord] ]

melted_cormat <- melt(cormat)
melted_cormat$diag <- melted_cormat$Var1 == melted_cormat$Var2
melted_cormat$diag[!melted_cormat$diag] <- NA
```

```{r fig.width=3.5, fig.height=2.5}
ggplot(data = melted_cormat, aes(x=Var2, y=Var1, fill=value)) + 
  geom_tile(color = "darkgrey")+
  geom_tile(data = melted_cormat[!is.na(melted_cormat$diag), ], aes(color = diag), size = 0.5) +
  scale_color_manual(guide = FALSE, values = c(`TRUE` = "black"))+
  scale_fill_gradient2(high = "#FDE725FF", low = "#440154FF", mid = "white", midpoint = 0, limit = c(-1,1), space = "Lab", name="Spearman\nCorrelation") +
  geom_text(aes(Var2, Var1, label = value), color = "black", size = 3) +
theme(axis.title = element_blank(), 
  panel.background = element_blank(), axis.text.x = element_text(angle=45, hjust=1,vjust=1), axis.text=element_text(color='black', size=9))
```
# Model with 9 factors 

```{r}
Y <- as.data.frame(scale(metadata, center = FALSE, scale = TRUE))
Y$added_col <- 1
C <- pinv(as.matrix(Y)) %*% as.matrix(X)
Xpred <- as.matrix(Y) %*% C
```

```{r}
C <- C[1:9,]
rownames(C) <- colnames(metadata)
```

```{r}
error_matrix <- abs(Xpred-X)
sd <- apply(X,2,sd) #sd for each site
errormat_by_sd <- sweep(error_matrix, 2, sd, FUN = '/')
errormat_by_sd_mean <- apply(errormat_by_sd,2,mean)
```

```{r}
hist(errormat_by_sd_mean)
```
```{r}
keep_sites <- as.data.frame(errormat_by_sd_mean[errormat_by_sd_mean < 0.5])
dim(keep_sites)
```
# Check p value of the correlation between Mpred and Mo for these sites

```{r}
Mo_filt <- X[, rownames(keep_sites)]
Mpred_filt <- as.data.frame(Xpred[, rownames(keep_sites)])

corr_matrix <- as.data.frame(mapply(cor,Mo_filt,Mpred_filt))

pvalues <- as.data.frame(t(mapply(\(x, y) cor.test(x, y)[c('estimate', 'p.value')], Mo_filt, Mpred_filt)))

pvalues$estimate <- as.numeric(pvalues$estimate)
pvalues$p.value <- as.numeric(pvalues$p.value)
pvalues$FDR <- p.adjust(pvalues$p.value, method = "BH")
```

```{r fig.width=3, fig.height=2}
ggplot(pvalues, aes(FDR)) + geom_histogram(fill="lightblue", color='black') + theme_classic() +
  ylab('Count') + xlab('FDR') +
  theme(axis.text = element_text(size=8), axis.title = element_text(size=10), 
        panel.border = element_rect(fill=NA, color='black', linewidth = 0.1))
```

```{r fig.width=3, fig.height=2}
keep_sites <- pvalues[pvalues$FDR < 0.01,]

ggplot(keep_sites, aes(FDR)) + geom_histogram(fill="lightblue", color='black') + theme_classic() +
  ylab('Count') + xlab('Adjusted pvalue') +
  theme(axis.text = element_text(size=8), axis.title = element_text(size=10), 
        panel.border = element_rect(fill=NA, color='black', linewidth = 0.1))
```
```{r}
dim(keep_sites)
```
```{r}
C_filtered <- C[, rownames(keep_sites)]
rownames(C_filtered) <- colnames(metadata)
```

```{r fig.width=3, fig.height=2.7}
C_filtered_norm <- t(C_filtered)
C_filtered_norm <- normalize.rows(as.matrix(C_filtered_norm), method = "euclidean")
C_filtered_norm <-C_filtered_norm[, c(9,3,1,2,4,5,6,7,8)]
colnames(C_filtered_norm) <- c("% Epithelial cells", "% Blood Neutrophils","Age","Sex","Hemoglobin","RDW","Urea","HDL","Batch")
```

```{r fig.width=3.3, fig.height=3.7}
set.seed(100)
pheatmap(C_filtered_norm, cluster_rows = T, cluster_cols = F, annotation_names_col = T, annotation_names_row=F, clustering_method = "complete", fontsize_col = 9,fontsize_row = 0.1)
```

```{r}
C_final <- round(C_filtered_norm, 3) %>% as.data.frame()
```

# Linear models -> pvalue

```{r}
df_final <- merge(metadata,X[, rownames(keep_sites)], by=0)
rownames(df_final) <- df_final$Row.names
df_final$Row.names <- NULL
```

```{r}
linear_models <- purrr::map(
  df_final[,-c(1:9)],  # set the elements to iterate over: only sites
  ~lm(.x ~ Age+Sex+X..Blood.Neutrophils+Hemoglobin+RDW+Urea+HDL+X..Epithelial.cells+Batch, data = df_final))
```

```{r}
linear_models2 <- map_df(linear_models, broom::tidy, .id = 'variable') %>% as.data.frame()
linear_models2$FDR <- p.adjust(linear_models2$p.value, method = "BH")

linear_models2 <- linear_models2[!linear_models2$term == '(Intercept)', ]
linear_models2 <- linear_models2[, -c(3:6)]

linear_models3 <- linear_models2 %>%
  pivot_wider(names_from = term, values_from = FDR) %>% as.data.frame()

rownames(linear_models3) <- linear_models3$variable
linear_models3$variable <- NULL
colnames(linear_models3) <- paste0('FDR','_', colnames(linear_models3))
linear_models3 <- round(linear_models3, 3)
```

# Keep sites with FDR < 0.1, then for each site rank factors and keep only sites in which the factor of interest is in the first rank

```{r}
abs_matrix <- abs(C_final)

sex_specific <- rownames(linear_models3[linear_models3$FDR_Sex < 0.1, ])
rank_sex <- apply(abs_matrix[sex_specific, ], 1, rank) %>% as.data.frame()
sex_specific2 <- names(rank_sex)[which(rank_sex['Sex',] == 9, arr.ind=T)[, "col"]]

epi_specific <- rownames(linear_models3[linear_models3$FDR_X..Epithelial.cells < 0.1, ])
rank_epi <- apply(abs_matrix[epi_specific, ], 1, rank) %>% as.data.frame()
epi_specific2 <- names(rank_epi)[which(rank_epi['% Epithelial cells',] == 9, arr.ind=T)[, "col"]]

age_specific <- rownames(linear_models3[linear_models3$FDR_Age < 0.1, ])
rank_age <- apply(abs_matrix[age_specific, ], 1, rank) %>% as.data.frame()
age_specific2 <- names(rank_age)[which(rank_age['Age',] == 9, arr.ind=T)[, "col"]]

batch_specific <- rownames(linear_models3[linear_models3$FDR_Batch < 0.1, ])
rank_batch <- apply(abs_matrix[batch_specific, ], 1, rank) %>% as.data.frame()
batch_specific2 <- names(rank_batch)[which(rank_batch['Batch',] == 9, arr.ind=T)[, "col"]]

neut_specific <- rownames(linear_models3[linear_models3$FDR_X..Blood.Neutrophils < 0.1, ])
rank_neut <- apply(abs_matrix[neut_specific, ], 1, rank) %>% as.data.frame()
neut_specific2 <- names(rank_neut)[which(rank_neut['% Blood Neutrophils',] == 9, arr.ind=T)[, "col"]]

hem_specific <- rownames(linear_models3[linear_models3$FDR_Hemoglobin < 0.1, ])
rank_hem <- apply(abs_matrix[hem_specific, ], 1, rank) %>% as.data.frame()
hem_specific2 <- names(rank_hem)[which(rank_hem['Hemoglobin',] == 9, arr.ind=T)[, "col"]]

rdw_specific <- rownames(linear_models3[linear_models3$FDR_RDW < 0.1, ])
rank_rdw <- apply(abs_matrix[rdw_specific, ], 1, rank) %>% as.data.frame()
rdw_specific2 <- names(rank_rdw)[which(rank_rdw['RDW',] == 9, arr.ind=T)[, "col"]]

urea_specific <- rownames(linear_models3[linear_models3$FDR_Urea < 0.1, ])
rank_urea <- apply(abs_matrix[urea_specific, ], 1, rank) %>% as.data.frame()
urea_specific2 <- names(rank_urea)[which(rank_urea['Urea',] == 9, arr.ind=T)[, "col"]]

hdl_specific <- rownames(linear_models3[linear_models3$FDR_HDL < 0.1, ])
rank_hdl <- apply(abs_matrix[hdl_specific, ], 1, rank) %>% as.data.frame()
hdl_specific2 <- names(rank_hdl)[which(rank_hdl['HDL',] == 9, arr.ind=T)[, "col"]]
```

```{r}
C_final2 <- C_final

C_final2$Sex_specific <- ifelse(rownames(C_final2) %in% sex_specific2, 'Yes', 'No')
C_final2$Age_specific <- ifelse(rownames(C_final2) %in% age_specific2, 'Yes', 'No')
C_final2$Batch_specific <- ifelse(rownames(C_final2) %in% batch_specific2, 'Yes', 'No')
C_final2$Epi_specific <- ifelse(rownames(C_final2) %in% epi_specific2, 'Yes', 'No')
C_final2$Neut_specific <- ifelse(rownames(C_final2) %in% neut_specific2, 'Yes', 'No')
C_final2$Hem_specific <- ifelse(rownames(C_final2) %in% hem_specific2, 'Yes', 'No')
C_final2$RDW_specific <- ifelse(rownames(C_final2) %in% rdw_specific2, 'Yes', 'No')
C_final2$Urea_specific <- ifelse(rownames(C_final2) %in% urea_specific2, 'Yes', 'No')
C_final2$Hdl_specific <- ifelse(rownames(C_final2) %in% hdl_specific2, 'Yes', 'No')
```

```{r}
age_pos <- rownames(C_final2[C_final2$Age_specific == 'Yes' & C_final2$Age > 0, ])
age_neg <- rownames(C_final2[C_final2$Age_specific == 'Yes' & C_final2$Age < 0, ])

epi_pos <- rownames(C_final2[C_final2$Epi_specific == 'Yes' & C_final2$`% Epithelial cells` > 0, ])
epi_neg <- rownames(C_final2[C_final2$Epi_specific == 'Yes' & C_final2$`% Epithelial cells` < 0, ])

neut_pos <- rownames(C_final2[C_final2$Neut_specific == 'Yes' & C_final2$`% Blood Neutrophils` > 0, ])
neut_neg <- rownames(C_final2[C_final2$Neut_specific == 'Yes' & C_final2$`% Blood Neutrophils` < 0, ])

hem_pos <- rownames(C_final2[C_final2$Hem_specific == 'Yes' & C_final2$Hemoglobin > 0, ])
hem_neg <- rownames(C_final2[C_final2$Hem_specific == 'Yes' & C_final2$Hemoglobin < 0, ])

rdw_pos <- rownames(C_final2[C_final2$RDW_specific == 'Yes' & C_final2$RDW > 0, ])
rdw_neg <- rownames(C_final2[C_final2$RDW_specific == 'Yes' & C_final2$RDW < 0, ])

urea_pos <- rownames(C_final2[C_final2$Urea_specific == 'Yes' & C_final2$Urea > 0, ])
urea_neg <- rownames(C_final2[C_final2$Urea_specific == 'Yes' & C_final2$Urea < 0, ])

hdl_pos <- rownames(C_final2[C_final2$Hdl_specific == 'Yes' & C_final2$HDL > 0, ])
hdl_neg <- rownames(C_final2[C_final2$Hdl_specific == 'Yes' & C_final2$HDL < 0, ])

batch_pos <- rownames(C_final2[C_final2$Batch_specific == 'Yes' & C_final2$Batch > 0, ])
batch_neg <- rownames(C_final2[C_final2$Batch_specific == 'Yes' & C_final2$Batch < 0, ])
```

# Some heatmaps

```{r fig.width=3, fig.height=3}
set.seed(100)
breaksList = seq(-1, 1, by = 0.1)
pheatmap(C_final[sex_specific2, ], cluster_rows = F, cluster_cols = F, annotation_names_col = T, annotation_names_row=F, clustering_method = "complete", fontsize_col = 7,fontsize_row = 8,
         breaks = breaksList,color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaksList)))
```
```{r fig.width=1.4, fig.height=2.5}
set.seed(100)
pheatmap(C_final[c(age_neg, age_pos), ], cluster_rows = F, cluster_cols = F, annotation_names_col = T, annotation_names_row=F, clustering_method = "complete", fontsize_col = 8,fontsize_row = 0.001,
         breaks = breaksList,color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaksList)))
```
# Some plots

## Chr X genes

```{r fig.width=1.7, fig.height=1.5}
sex_df <- df_final[, c('Sex', sex_specific2)]
sex_df$Sex <- ifelse(sex_df$Sex == 0,'M','F')
sex_df$Sex <- factor(sex_df$Sex, levels=c('M','F'))

sex_df <- sex_df %>% group_by(Sex) %>% 
  summarise(across(everything(), mean)) 
rownames(sex_df) <- sex_df$Sex
sex_df$Sex <- NULL
sex_df <- t(sex_df)
colnames(sex_df) <- c('M','F')

set.seed(100)
pheatmap(sex_df, cluster_rows = T, cluster_cols = F, annotation_names_col = T, annotation_names_row=F, clustering_method = "complete", fontsize_col = 8,fontsize_row = 7, angle_col = "0")
```
